---
title: "Traffic Analysis R Analysis"
output: html_document
date: "2025-04-08"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
#Reading in the data
traffic2023 <- read.csv("/Users/kylelevesque/Downloads/data 2/ct_traffic_data_2023.csv")
head(traffic2023)
dim(traffic2023)
summary(traffic2023$aadt)

```


```{r}
#Just getting a feel for the data
length(unique(traffic2023$route_id))
sum(duplicated(traffic2023$route_id))
#There are some duplicated routes so can't really so can't really analyze by route_id but instead coordinates

```

```{r}
#This code just gets the latitude and longitude into coordinates
traffic2023$from <- paste(traffic2023$slat, traffic2023$slon, sep = ",")
traffic2023$to <- paste(traffic2023$elat, traffic2023$elon, sep = ",")

#This gets the amount of traffic flow from one specific location to another specific location
traffic_flow <- aggregate(aadt ~ from + to, data = traffic2023, sum)

#This is the total amount of traffic flow from one specific location to any location 
total_outbound <- aggregate(aadt ~ from, data = traffic_flow, sum)

#This merges the total_outbound data with the traffic_flow data 
traffic_flow <- merge(traffic_flow, total_outbound, by = "from", suffixes = c("_to", "_total"))

traffic_flow$turn_prob <- traffic_flow$aadt_to / traffic_flow$aadt_total

# Step 5: View results
head(traffic_flow)
```


```{r}
#Get all unique "nodes". This is takes all of the possible coordinate combinations from the from and to columns and gets the unique locations.  
nodes <- unique(c(traffic_flow$from, traffic_flow$to))
length(nodes)
dim(traffic_flow)
#This creates an empty stochastic matrix with the nodes
n <- length(nodes)
P <- matrix(0, nrow = n, ncol = n)
rownames(P) <- nodes
colnames(P) <- nodes
#This fills in the matrix with the corresponding turn probability data 
for (i in 1:nrow(traffic_flow)) {
  from_node <- traffic_flow$from[i]
  to_node <- traffic_flow$to[i]
  prob <- traffic_flow$turn_prob[i]
  P[from_node, to_node] <- prob
}

#This gets the row sums to make sure that they add up to 1 
row_sums <- rowSums(P)
summary(row_sums)

#There appears to be some dead ends as the mean across rows is not 1. Therefore, there must be some locations that do not go anywhere. 
dead_ends <- rownames(P)[rowSums(P) == 0]
length(dead_ends)
#There appears to be 1090 dead ends out of 5333 intersections, which could be a little high 
#write.csv(P, file = "/Users/kylelevesque/Downloads/stochastic_matrix.csv", row.names = TRUE)
```

```{r}
dead_end_flag <- as.numeric(rowSums(P) == 0)
P_with_col <- cbind(P, is_dead_end = dead_end_flag)
dim(P_with_col)
new_row <- rep(0, ncol(P_with_col))
P_augmented <- rbind(P_with_col, new_row)
rownames(P_augmented)[nrow(P_augmented)] <- "dead_end"
P_augmented[nrow(P_augmented), ncol(P_augmented)] <- 1
summary(rowSums(P_augmented))
dim(P_augmented)
#write.csv(P_augmented, file = "/Users/kylelevesque/Downloads/MATH 222/stochastic_matrix2.csv", row.names = T)
```
```{r}
#Markov Chain/Steady State
eig <- eigen(t(P_augmented))
idx <- which.max(Re(eig$values))
steady_state <- Re(eig$vectors[, idx])
steady_state <- steady_state / sum(steady_state)
head(steady_state)
sum(steady_state)


```

```{r}
node <- rownames(P_augmented)
steady_state_df <- data.frame(
  node = node,
  steady_state_prob = steady_state
)
#write.csv(steady_state_df, file = "/Users/kylelevesque/Downloads/MATH 222/steady_state_df.csv", row.names = T)
steady_state_df <- steady_state_df[order(-steady_state_df$steady_state), ]
steady_state_df$lon <- as.numeric(sub(",.*", "", steady_state_df$node))
steady_state_df$lat <- as.numeric(sub(".*,", "", steady_state_df$node))


```

```{r}
library(maps)
library(mapproj)
```


```{r}
# Plot a base map of Connecticut
png("connecticut_top10_map.png", width = 1200, height = 900)
map("county", "connecticut", col = "white", fill = TRUE)
title("Top 10 Busiest Intersections in Connecticut")
points(jitter(steady_state_df$lon[1:10], amount = 0.005), jitter(steady_state_df$lat[1:10], amount = 0.005), col = rainbow(10), pch = 19, cex = 2)
dev.off()
```


```{r}
library(dplyr)
node_route_df <- traffic2023 %>%
  select(route_id, from, to) %>%
  tidyr::pivot_longer(cols = c(from, to), names_to = "type", values_to = "node") %>%
  distinct()
node_route_df
```

```{r}
steady_state_with_routes <- merge(steady_state_df, node_route_df, by = "node", all.x = TRUE)
top10_with_routes <- merge(steady_state_df[1:10, ], steady_state_with_routes[, c("node", "route_id")], by = "node")
top10_with_routes_unique <- top10_with_routes[!duplicated(top10_with_routes$node), ]

```

```{r}
png("CT_busiest_intersections.png", width = 1200, height = 900)

map("county", "connecticut", col = "lightgray", fill = TRUE, bg = "white")
title("Top 10 Busiest Routes in Connecticut in the Long Term")

# Create a color palette for route_ids
# Generate a unique color for each route_id
unique_route_ids <- unique(top10_with_routes$route_id)
colors <- rainbow(length(unique_route_ids))

# Assign colors to the routes in your dataframe
top10_with_routes$color <- colors[match(top10_with_routes$route_id, unique_route_ids)]

# Plot the top 10 busiest intersections with the assigned colors
points(jitter(top10_with_routes$lon, amount = 0.005), jitter(top10_with_routes$lat, amount = 0.005), col = top10_with_routes$color, pch = 19, cex = 1.5)

# Add a legend with route_id and the corresponding color
legend("topleft", 
       legend = unique_route_ids, 
       fill = colors, 
       title = "Route ID", 
       cex = 0.8)

dev.off()
```

```{r}
#Start at a random road and see probability of ending up at different nodes
dead_ends <- rowSums(P) == 0
valid_nodes <- rownames(P)[!dead_ends]

set.seed(500)  # for reproducibility
start_node <- sample(valid_nodes, 1)
start_index <- which(rownames(P_augmented) == start_node)
```

```{r}
v <- rep(0, nrow(P_augmented))
v[start_index] <- 1
```

```{r}
steps <- 3
v_current <- v

for (i in 1:steps) {
  v_current <- v_current %*% P_augmented
}
```

```{r}
results <- data.frame(
  node = rownames(P_augmented),
  probability = as.vector(v_current)
)

results_sorted <- results[order(-results$probability), ]
random_prob_results <- results_sorted[results_sorted$probability != 0, ]

random_prob_results$lon <- as.numeric(sub(",.*", "", random_prob_results$node))
random_prob_results$lat <- as.numeric(sub(".*,", "", random_prob_results$node))

```
```{r}
from_matches <- merge(random_prob_results, traffic2023[, c("route_id", "from")],
                      by.x = "node", by.y = "from", all.x = TRUE)

# Match to_node
to_matches <- merge(random_prob_results, traffic2023[, c("route_id", "to")],
                    by.x = "node", by.y = "to", all.x = TRUE)

# Combine both matches
combined_matches <- rbind(from_matches, to_matches)

# Remove duplicates (some nodes may appear in both from and to)
random_path <- combined_matches[c(1, 2, 7, 8), ]

#random_path <- write.csv(random_path, file = "/Users/kylelevesque/Downloads/MATH 222/random_path.csv", row.names = T)

```


```{r}
unique_route <- unique(random_path$route_id[1:3])
colors <- rainbow(length(unique_route_ids))

random_path$color <- colors[match(random_path$route_id, unique_route)]

start_lon <- as.numeric(sub(",.*", "", start_node))
start_lat <- as.numeric(sub(".*,", "", start_node))
```


```{r}
png("MostProbableDestinations.png", width = 600, height = 800)
library(maps)

map("county", "connecticut,fairfield", col = "lightblue", fill = TRUE, lwd = 0.5)
title("Most Probable Destinations after 3 Turns")

points(random_path$lon[1:3], random_path$lat[1:3], pch = 19, col = random_path$color, cex = 2)
text(random_path$lon, random_path$lat, 
     labels = round(random_path$probability, 3), pos = 4, cex = 0.7, col = "black")

points(start_lon, start_lat, pch = 8, col = "orange", cex = 2)
text(start_lon, start_lat, labels = "Start", pos = 2, cex = 1, col = "orange")

legend("topright", 
       legend = unique_route, 
       fill = colors, 
       title = "Route ID", 
       cex = 0.8)

legend("bottomright", 
       legend = paste("Dead End Probability:", round(random_path$probability[4], 3)),
       bty = "o",
       cex = 0.8,
       text.col = "gray20",
       box.col = "gray40",
       bg = "white",
       inset = c(0.02, 0.02)
)

dev.off()
```


1. Quick statistics: average daily flow

2. Long term probability, multiply matrix a ton of times: good

Have steady_state vector for matrix. Probability that a car will be at a certain intersection in the long run. Have the busiest intersections based on steady state probabilities. 

3. Starting at a random road, what is the probability that will end up on another road: good 
Have




